---
title: "Uncovering Player Archetypes in Global Football"
author: "Kevin Davies"
date: "`r Sys.Date()`"
output: html_notebook
---

# Executive Summary

This notebook presents a full data analysis pipeline applied to the FIFA 21 player
dataset. The central research question is: **Can we predict a player's weekly wage
from their six skill attributes and player type?**

The analysis follows six stages:

1. **Data Acquisition** — loading the FIFA 21 dataset directly from a published URL
2. **Data Understanding** — exploring how player scores are constructed
3. **Data Cleaning** — systematic framework covering irrelevant data, missing values,
   variable types, numerical domains, duplicates, outliers, and consistency checks
4. **Exploratory Data Analysis** — visualising distributions and relationships
5. **Statistical Modelling** — multiple linear regression to predict player wage
6. **Dimension Reduction** — PCA to identify underlying skill dimensions

**Key findings:**
- Dribbling is the strongest predictor of wage among the six skill scores
- Player type (Defender/Forward) commands a ~50% wage premium over Midfielders
  with identical skill scores
- The model explains 44.2% of wage variation
- PCA reveals two fundamental skill dimensions: technical attacking ability (PC1)
  and defensive/physical ability (PC2)

**Reproducibility note:** This notebook is fully self-contained. All data is loaded
directly from a public URL. Running all chunks from top to bottom in order will
reproduce all results exactly. A fixed random seed (set.seed(123)) is used for
the PCA downsampling step to ensure identical results on every run.

---

# 1. Data Acquisition

```{r message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(corrplot)

# Load raw dataset directly from Google Sheets published as CSV
# No local file required — data loads automatically from URL
soccer_raw <- read.csv(
  "https://docs.google.com/spreadsheets/d/e/2PACX-1vRWpw5hL7DUtcb9k7NsVNsE4BPV9QPZ-w1uBTovoE6TbHDur3a0dwSCmiRjcZlyX-lQ3_PhfnkFfBhl/pub?gid=0&single=true&output=csv"
)

# Confirm successful load
cat("Rows:", nrow(soccer_raw), "\n")
cat("Columns:", ncol(soccer_raw), "\n")
head(soccer_raw, 5)
```

The raw dataset contains 18,944 players across 106 columns, covering identity
information, skill scores, sub-attributes, and market data. Each row represents
one player.

---

# 2. Data Understanding

## 2.1 What is the FIFA 21 Dataset?

The FIFA 21 dataset contains player-level data from EA Sports' FIFA 21 video game.
Each row represents one player. The key attributes fall into a few groups:

- **Identity**: name, nationality, club, age, position
- **Overall & Potential**: composite scores (0–100) representing current ability and ceiling
- **Physical**: height, weight
- **6 Summary Skill Scores** (0–100 each): Pace, Shooting, Passing, Dribbling, Defending, Physicality
- **Detailed Sub-attributes (~30)**: e.g., `attacking_crossing`, `skill_dribbling`, `movement_sprint_speed`
- **Market info**: value (EUR), wage (EUR)

## 2.2 How Are Player Scores Calculated?

FIFA 21 player ratings are not arbitrary numbers — they are the result of a structured,
multi-stage process that transforms real world observations into a precise numerical scale.

**Stage 1 — Data Collection**
EA Sports employs over 6,000 volunteer scouts worldwide — coaches, professional scouts,
and football analysts — who watch real matches and assess players across more than 35
individual attributes.

**Stage 2 — Transforming Raw Observations into 0–100**
Take speed as a concrete example. A player's pace rating is anchored to a real athletic
benchmark — the 100 metre sprint. A pace rating of 99 corresponds to approximately 9.8
seconds for 100 metres — essentially world class sprinting speed. EA and their data
partners use GPS tracking technology and match data to measure real sprint speeds during
training and games, which are then converted to an equivalent 100 metre benchmark and
placed on the 0 to 100 scale. The same principle applies to every other attribute —
each one has a real world benchmark behind it.

**Stage 3 — Combining Attributes into Summary Scores**
Those individual attribute scores are then combined into the six summary scores.
Each summary score is a weighted average of its underlying sub-attributes:

```{r}
# FIFA 21 score breakdown — documented from EA's attribute system
score_breakdown <- tribble(
  ~Summary_Score,  ~Key_Sub_Attributes,
  "Pace",          "sprint_speed, acceleration",
  "Shooting",      "finishing, shot_power, long_shots, volleys, penalties, heading_accuracy",
  "Passing",       "short_passing, long_passing, vision, crossing, curve, free_kick_accuracy",
  "Dribbling",     "dribbling, ball_control, agility, balance, reactions",
  "Defending",     "standing_tackle, sliding_tackle, interceptions, defensive_awareness, heading_accuracy",
  "Physicality",   "stamina, strength, jumping, aggression"
)

score_breakdown
```

**Stage 4 — The Position-Weighted Overall**
The Overall rating is a position-weighted blend of the six summary scores, using
position-specific coefficients that add up to 1. A striker's Overall weights Pace
and Shooting most heavily. A central midfielder's Overall weights Passing and
Dribbling. A centre-back's Overall weights Defending and Physicality.

This means two players with the same Overall rating can have completely different
skill profiles — which is exactly why looking at underlying archetypes is more
informative than Overall rating alone.

## 2.3 Variable Overview

```{r}
# Group all column names by attribute type to understand dataset structure
names(soccer_raw) %>%
  as_tibble() %>%
  mutate(group = case_when(
    grepl("^attacking_", value) ~ "Attacking sub-attrs",
    grepl("^skill_", value)     ~ "Skill sub-attrs",
    grepl("^movement_", value)  ~ "Movement sub-attrs",
    grepl("^power_", value)     ~ "Power sub-attrs",
    grepl("^mentality_", value) ~ "Mentality sub-attrs",
    grepl("^defending_", value) ~ "Defending sub-attrs",
    grepl("^gk_|^goalkeeping_", value) ~ "Goalkeeper attrs",
    TRUE ~ "Identity / Summary"
  )) %>%
  count(group, name = "num_columns") %>%
  arrange(desc(num_columns))
```

The dataset contains approximately 30 detailed sub-attributes grouped into
attacking, skill, movement, power, mentality, and defending categories —
providing a rich basis for identifying player archetypes.

---

# 3. Data Cleaning

We follow a systematic framework covering eight categories of data quality checks:

| # | Issue Type          | What we check                              |
|---|---------------------|--------------------------------------------|
| 1 | Irrelevant rows     | Goalkeepers removed — different attr space |
| 2 | Irrelevant columns  | Administrative and redundant columns       |
| 3 | Missing values      | NAs across all columns                     |
| 4 | Variable types      | Correct data types for each column         |
| 5 | Numerical domains   | Values within expected ranges              |
| 6 | Duplicate rows      | Players recorded more than once            |
| 7 | Outliers            | Extreme values in key numeric columns      |
| 8 | Consistency         | Logical relationships between variables    |

## 3.1 Remove Goalkeepers

```{r}
cat("Players with GK position:", sum(grepl("GK", soccer_raw$player_positions)), "\n")

soccer <- soccer_raw %>%
  filter(!grepl("GK", player_positions))

cat("Remaining outfield players:", nrow(soccer), "\n")
```

**Justification:** Goalkeepers are excluded because they operate in a fundamentally
different attribute space from outfield players. Their skill profile is dominated
by goalkeeping-specific attributes — reflexes, diving, handling, kicking — that
do not exist for outfield players. Including them would distort any clustering or
modelling of outfield archetypes. They are not simply another archetype — their
attribute structure is incomparable to outfield players.

## 3.2 Remove Goalkeeper-Specific Columns

```{r}
# Two passes to ensure complete removal of all GK columns
soccer <- soccer %>%
  select(-matches("^gk_|^goalkeeping_")) %>%
  select(-any_of(c(
    "gk_diving", "gk_handling", "gk_kicking",
    "gk_reflexes", "gk_speed", "gk_positioning"
  )))

cat("Columns after removing GK attrs:", ncol(soccer), "\n")
```

**Justification:** Goalkeeper-specific attribute columns contain no meaningful
information for outfield players — they are either structurally empty or filled
with zeros. Retaining them would introduce noise into the analysis without adding
any useful signal.

## 3.3 Remove Irrelevant Columns

```{r}
soccer <- soccer %>%
  select(-any_of(c(
    # Database identifiers — not player attributes
    "sofifa_id", "player_url",
    # Redundant identity — short_name and age are retained
    "long_name", "dob",
    # Contextual metadata — not skill attributes
    "nationality", "club_name", "league_name", "league_rank",
    # Financial — value_eur removed; wage_eur intentionally KEPT as target variable
    "value_eur",
    # Categorical descriptors — not numerical skill scores
    "preferred_foot", "work_rate", "body_type", "real_face",
    # Mostly missing (~40% NA) — not relevant to skill analysis
    "release_clause_eur",
    # Administrative metadata — not player skill attributes
    "player_tags", "team_position", "loaned_from", "joined",
    "nation_position", "nation_jersey_number", "team_jersey_number",
    "contract_valid_until",
    # Not direct skill measurements
    "international_reputation", "player_traits",
    # Completely empty — deprecated in FIFA 21, replaced by defensive_awareness
    "defending_marking",
    # Position rating columns — derived from skill attributes, therefore redundant
    "ls", "st", "rs", "lw", "lf", "cf", "rf", "rw",
    "lam", "cam", "ram", "lm", "lcm", "cm", "rcm", "rm",
    "lwb", "ldm", "cdm", "rdm", "rwb", "lb", "lcb", "cb", "rcb", "rb"
  )))

cat("Columns after removing irrelevant vars:", ncol(soccer), "\n")
```

**Justification summary:**
- Administrative columns (IDs, URLs, jersey numbers, contract dates) provide no
  information about player skill and are irrelevant to our analysis questions.
- `value_eur` is a market valuation influenced by many non-skill factors.
  `wage_eur` is retained as our target variable for modelling.
- `release_clause_eur` was mostly missing (~40% NA) and not relevant to skill analysis.
- Position rating columns (e.g., `st`, `cam`, `cb`) are computed directly from the
  skill sub-attributes already retained — keeping them would be redundant.
- `defending_marking` was deprecated by EA Sports in FIFA 21 and contains no data.

## 3.4 Check for Missing Values

```{r}
# Identify any columns with missing values
na_summary <- colSums(is.na(soccer))
na_summary[na_summary > 0] %>%
  sort(decreasing = TRUE) %>%
  as_tibble(rownames = "column") %>%
  rename(missing = value)
```

```{r}
cat("Total missing values:", sum(is.na(soccer)), "\n")
cat("Total cells:", nrow(soccer) * ncol(soccer), "\n")
cat("Missing %:", round(100 * sum(is.na(soccer)) /
      (nrow(soccer) * ncol(soccer)), 2), "%\n")
```

## 3.5 Select Analysis Variables

```{r}
# Identity and target variable columns
# wage_eur is retained as the target variable — NOT treated as a predictor
# age is retained as it may independently influence wage
soccer_context <- soccer %>%
  select(short_name, player_positions, wage_eur, age)

# Skill attributes — these will be the predictors in modelling
soccer_style <- soccer %>%
  select(
    overall, potential,
    pace, shooting, passing, dribbling, defending, physic,
    height_cm, weight_kg,
    weak_foot, skill_moves,
    starts_with("attacking_"),
    starts_with("skill_"),
    starts_with("movement_"),
    starts_with("power_"),
    starts_with("mentality_"),
    starts_with("defending_")
  )

# Combine into one clean analysis dataset
soccer_clean <- bind_cols(soccer_context, soccer_style)

cat("Final dataset:", nrow(soccer_clean), "players,", ncol(soccer_clean), "columns\n")
```

## 3.6 Check Variable Types

```{r}
# Verify data types — skill attributes should be numeric (int or num)
# Identity columns (short_name, player_positions) will be chr — expected
str(soccer_clean)
```

**Finding:** All skill attribute columns are stored as integers — correct for
numerical analysis. Identity columns are character type as expected.

## 3.7 Check Numerical Domains

```{r}
# Verify all numeric columns fall within expected ranges
soccer_clean %>%
  select(where(is.numeric)) %>%
  summary()
```

```{r}
# Flag zero wages — not realistic for a registered professional footballer
cat("Players with zero wage:", sum(soccer_clean$wage_eur == 0), "\n")

# Remove zero wage players
# Zero wage is not realistic and likely represents missing or incorrectly recorded data
# Since wage_eur is our target variable, rows with zero wage cannot be used in modelling
soccer_clean <- soccer_clean %>%
  filter(wage_eur > 0)

cat("Players remaining after removing zero wage:", nrow(soccer_clean), "\n")
```

**Finding:** All skill scores fall within the expected 0–100 range. Height ranges
from 155cm to 203cm and weight from 50kg to 110kg — both realistic. 195 players
had a recorded wage of zero — removed as likely data recording errors.

## 3.8 Check for Duplicate Rows

```{r}
duplicate_count <- sum(duplicated(soccer_clean))
cat("Duplicate rows:", duplicate_count, "\n")
```

**Finding:** No duplicate rows found. Every row represents a unique player.

## 3.9 Check for Outliers

```{r fig.width=7, fig.height=4}
ggplot(soccer_clean, aes(y = wage_eur)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Distribution of Player Wages (log scale)",
       y = "Weekly Wage (EUR)") +
  theme_minimal()
```

```{r}
# Top 10 highest earning players
soccer_clean %>%
  arrange(desc(wage_eur)) %>%
  select(short_name, player_positions, overall, wage_eur) %>%
  head(10)
```

**Finding:** Wage is heavily right skewed. The top earners are world class players
whose wages are legitimate real world values — not data errors. These genuine
outliers reflect real world wage inequality in football and are retained.

## 3.10 Consistency Check

```{r}
# A player's Overall should never exceed their Potential
inconsistent <- sum(soccer_clean$overall > soccer_clean$potential)
cat("Players where Overall exceeds Potential:", inconsistent, "\n")
```

**Finding:** No players have an Overall rating exceeding their Potential — the
data is internally consistent on this logical relationship.

```{r}
cat("Final clean dataset:", nrow(soccer_clean), "players,",
    ncol(soccer_clean), "columns\n")
```

---

# 4. Exploratory Data Analysis (EDA)

## 4.1 Distribution of Overall Ratings

```{r fig.width=8, fig.height=4}
ggplot(soccer_clean, aes(x = overall)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Overall Ratings — FIFA 21 Outfield Players",
    x = "Overall Rating", y = "Number of Players"
  ) +
  theme_minimal()
```

The distribution is approximately normal, with most players clustering between
60 and 70. Truly elite players rated 85 and above are rare, reflecting the real
world scarcity of world class talent.

## 4.2 Messi's Six Skill Scores — A Concrete Example

```{r fig.width=8, fig.height=4}
messi <- soccer_clean %>%
  filter(short_name == "L. Messi") %>%
  select(short_name, pace, shooting, passing, dribbling, defending, physic) %>%
  pivot_longer(-short_name, names_to = "skill", values_to = "score")

ggplot(messi, aes(x = reorder(skill, score), y = score, fill = skill)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = score), hjust = -0.3, size = 4) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 110)) +
  labs(
    title = "Messi's Six Skill Scores — FIFA 21",
    subtitle = "Overall: 93 | Position: RW, ST, CF",
    x = "Skill", y = "Score"
  ) +
  theme_minimal()
```

Messi's Overall of 93 is built primarily on elite Dribbling (95), Shooting (92),
and Passing (91). His Defending score of 38 is very low — yet his Overall remains
93 because his position weights attacking skills heavily. This illustrates why
Overall alone is insufficient to describe a player's full skill profile.

## 4.3 Distributions of the 6 Summary Skill Scores

```{r fig.width=10, fig.height=6}
soccer_clean %>%
  select(pace, shooting, passing, dribbling, defending, physic) %>%
  pivot_longer(everything(), names_to = "skill", values_to = "score") %>%
  ggplot(aes(x = score, fill = skill)) +
  geom_histogram(binwidth = 3, color = "white", show.legend = FALSE) +
  facet_wrap(~skill, ncol = 3) +
  labs(title = "Distributions of the 6 FIFA 21 Skill Scores",
       x = "Score", y = "Count") +
  theme_minimal()
```

Defending shows a bimodal distribution reflecting the attacker-defender divide.
Shooting is widely spread suggesting high variability in attacking ability.
Dribbling and Physicality are more normally distributed, while Pace is left skewed.

## 4.4 Boxplots by Broad Position Group

```{r fig.width=10, fig.height=5}
soccer_eda <- soccer_clean %>%
  mutate(position_group = case_when(
    grepl("CB|LB|RB|LWB|RWB", player_positions) ~ "Defender",
    grepl("CDM|CM|CAM|LM|RM", player_positions)  ~ "Midfielder",
    grepl("ST|CF|LW|RW|LF|RF", player_positions) ~ "Forward",
    TRUE ~ "Other"
  )) %>%
  filter(position_group != "Other")

soccer_eda %>%
  select(position_group, pace, shooting, passing, dribbling, defending, physic) %>%
  pivot_longer(-position_group, names_to = "skill", values_to = "score") %>%
  ggplot(aes(x = position_group, y = score, fill = position_group)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  facet_wrap(~skill, ncol = 3) +
  labs(title = "Skill Score Distributions by Position Group",
       x = "Position Group", y = "Score") +
  theme_minimal()
```

Clear position-based patterns emerge. Defending shows the biggest separation —
defenders score around 65 on average while forwards score as low as 25–30.
Forwards are highest on Shooting. Midfielders are the most balanced group.
Pace is similar across all three groups suggesting speed is a universal requirement.

## 4.5 Correlation Matrix of Skill Scores

```{r fig.width=7, fig.height=6}
skill_cols <- soccer_clean %>%
  select(overall, pace, shooting, passing, dribbling, defending, physic) %>%
  drop_na()

cor_matrix <- cor(skill_cols)

corrplot(cor_matrix,
         method = "color", type = "upper",
         addCoef.col = "black", tl.col = "black",
         number.cex = 0.8,
         title = "Correlation Matrix of FIFA 21 Skill Scores",
         mar = c(0, 0, 2, 0))
```

The strongest positive relationship is between Passing and Dribbling (0.83).
Shooting and Defending show a negative correlation (–0.40), confirming the
attacker-defender divide. Shooting and Physicality (0.02) are essentially
uncorrelated — physical strength is an independent dimension. These widespread
correlations motivate the use of PCA to identify independent underlying dimensions.

---

# 5. Statistical Modelling — Predicting Player Wage

## 5.1 Research Question

**Can we predict a player's weekly wage from their six skill attributes and player type?**

We use **multiple linear regression** — a method that models the relationship between
a target variable and multiple predictors simultaneously. The target variable is
`wage_eur` — what we are trying to predict. The predictors are the six skill scores,
age, and player type.

Raw wage is heavily right skewed — the median is €3,000 but the maximum is €560,000.
A linear regression applied to raw wage would be distorted by extreme earners. We
therefore apply a **log transformation** to wage before modelling, which compresses
the scale and makes the distribution more suitable for linear regression.

## 5.2 Add Player Type as a Categorical Variable

```{r}
# Classify players into three position groups
# player_type is included as a categorical predictor — position may influence
# wage independently of skill scores
soccer_model <- soccer_clean %>%
  mutate(player_type = case_when(
    grepl("CB|LB|RB|LWB|RWB", player_positions) ~ "Defender",
    grepl("CDM|CM|CAM|LM|RM", player_positions)  ~ "Midfielder",
    grepl("ST|CF|LW|RW|LF|RF", player_positions) ~ "Forward",
    TRUE ~ "Other"
  )) %>%
  filter(player_type != "Other") %>%
  mutate(
    # Midfielder set as reference category — largest and most balanced group
    # All Defender and Forward coefficients interpreted relative to Midfielder
    player_type = factor(player_type, levels = c("Midfielder", "Defender", "Forward")),
    log_wage = log(wage_eur)
  )

cat("Players in model dataset:", nrow(soccer_model), "\n")
cat("Player type distribution:\n")
table(soccer_model$player_type)
```

Midfielder is set as the reference category — all Defender and Forward coefficients
are interpreted relative to a Midfielder with the same skill scores. Midfielder
was chosen as the reference because it is the largest group (7,094 players) and
the most balanced position in terms of skill attributes.

## 5.3 Visualise Log Wage Distribution

```{r fig.width=8, fig.height=4}
ggplot(soccer_model, aes(x = log_wage)) +
  geom_histogram(binwidth = 0.3, fill = "steelblue", color = "white") +
  scale_x_continuous(
    breaks = log(c(500, 1000, 2000, 3000, 10000, 50000, 200000)),
    labels = c("€500", "€1K", "€2K", "€3K", "€10K", "€50K", "€200K")
  ) +
  labs(
    title = "Distribution of Log-Transformed Weekly Wage",
    x = "Weekly Wage (EUR)", y = "Count"
  ) +
  theme_minimal()
```

The distribution remains bimodal after log transformation — two distinct peaks
around €500 and €2,000 per week — reflecting the two tier wage structure in
professional football. There is a real gap between €1,000 and €2,000 reflecting
the wage cliff between lower league and top league football.

## 5.4 Fit the Regression Model

```{r}
# Multiple linear regression
# Target: log(wage_eur)
# Predictors: six skill scores + age + player type (categorical)
model <- lm(log_wage ~ pace + shooting + passing + dribbling +
              defending + physic + age + player_type,
            data = soccer_model)

summary(model)
```

## 5.5 Interpret the Coefficients

```{r}
broom::tidy(model) %>%
  mutate(
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    p.value = round(p.value, 4),
    significant = ifelse(p.value < 0.05, "Yes", "No")
  ) %>%
  select(term, estimate, std.error, p.value, significant)
```

**Interpreting numerical predictors** — each coefficient tells us how much log wage
changes for a one unit increase in that skill score, holding all other predictors constant:

- **Dribbling (0.052):** The strongest skill predictor. Each one point increase in
  dribbling is associated with a 5.2% increase in log wage. On the original wage
  scale, a 10 point increase in dribbling corresponds to approximately e^(0.052×10) - 1
  = 68% higher weekly wage, all else equal.
- **Physicality (0.028):** Each one point increase is associated with a 2.8% increase
  in log wage. Physical players are well rewarded in the market.
- **Shooting (0.018), Defending (0.017), Passing (0.016):** All positive — better
  skilled players in these areas earn more, all else equal.
- **Pace (-0.007):** The only negative skill predictor. Pure pace alone, when all
  other skills are held constant, is associated with a slight decrease in log wage.
  Pace without supporting technical skills does not command a wage premium.
- **Age (-0.006):** Each additional year of age is associated with a 0.6% lower
  log wage when skill level is controlled for.

**Interpreting the categorical predictor — player type:**

- **Defender (+0.422):** Holding all skill scores and age constant, a Defender earns
  approximately e^0.422 - 1 = **52% more** than a Midfielder with identical attributes
  on the original wage scale.
- **Forward (+0.403):** A Forward earns approximately e^0.403 - 1 = **50% more** than
  a Midfielder with identical skill scores on the original wage scale.

Position itself commands a significant wage premium independent of skill level.

## 5.6 Overall Model Fit

```{r}
cat("R-squared:", round(summary(model)$r.squared, 4), "\n")
cat("Adjusted R-squared:", round(summary(model)$adj.r.squared, 4), "\n")
```

**R-squared: 0.442** — the model explains 44.2% of the variation in player wages.
The remaining 55.8% is driven by factors not in our model — club size, league,
commercial value, and contract negotiations.

**Adjusted R-squared: 0.4417** — almost identical to R-squared, confirming all
eight predictors are genuinely contributing and none are adding noise.

## 5.7 Check Model Assumptions

```{r fig.width=10, fig.height=8}
par(mfrow = c(2, 2))
plot(model)
```

**Residuals vs Fitted:** Slight curve in the red line — indicating some non-linearity.
The model makes systematic errors at certain wage levels.

**Normal Q-Q:** Points follow the diagonal line well in the middle but curve away
at both ends — heavy tails — meaning larger errors at extreme wage values.

**Scale-Location:** Red line slopes upward — the model is less accurate for higher
wage players, violating the constant variance assumption.

**Residuals vs Leverage:** Red line is flat and no points fall outside Cook's distance
boundaries — no single player is distorting the model.

**Overall assessment:** The model satisfies the leverage assumption but shows
violations of linearity and constant variance — consistent with the bimodal wage
structure. Recommendations: model the two wage tiers separately, or explore
mixture models to better account for this structure.

## 5.8 Visualise Predicted vs Actual Wage

```{r fig.width=8, fig.height=5}
soccer_model$predicted_log_wage <- predict(model, soccer_model)

ggplot(soccer_model, aes(x = predicted_log_wage, y = log_wage)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(
    title = "Predicted vs Actual Log Wage",
    x = "Predicted Log Wage", y = "Actual Log Wage"
  ) +
  theme_minimal()
```

Points close to the red diagonal line indicate accurate predictions. The model
performs well for average players but consistently underestimates elite earners.
The horizontal banding reflects the discrete nature of FIFA 21 wages recorded
in round numbers. This is consistent with the R-squared of 0.442.

---

# 6. Dimension Reduction — PCA

## 6.1 Why PCA?

Our dataset contains 28 skill sub-attributes — many of which are correlated with
each other. For example, Passing and Dribbling have a correlation of 0.83. When
many variables move together like this, it is difficult to identify what truly
and independently separates players from each other.

**Principal Component Analysis (PCA)** addresses this by:
- Taking all 28 correlated sub-attributes
- Finding new **uncorrelated** dimensions called principal components
- Each component captures a different independent source of variation
- A small number of components can represent most of the information

In short — PCA finds the underlying skill dimensions that truly separate players,
rather than looking at 28 correlated attributes individually. This connects directly
to the core goal of this project — uncovering player archetypes.

## 6.2 Prepare Data for PCA

```{r}
# Select only the detailed sub-attributes for PCA
# Summary scores (overall, pace etc.) excluded — derived from sub-attributes
# skill_moves excluded — it is a 1-5 star rating, not a continuous sub-attribute
# Wage, age, and identity columns excluded

# Downsample to create genuinely high dimensional data (more columns than rows)
# as recommended — 25 players, 28 sub-attribute columns
# set.seed ensures reproducibility — same 25 players on every run
set.seed(123)

pca_data <- soccer_clean %>%
  select(
    starts_with("attacking_"),
    starts_with("skill_"),
    starts_with("movement_"),
    starts_with("power_"),
    starts_with("mentality_"),
    starts_with("defending_")
  ) %>%
  select(-skill_moves) %>%
  drop_na() %>%
  sample_n(25)

cat("Rows:", nrow(pca_data), "\n")
cat("Columns:", ncol(pca_data), "\n")
```

The dataset is downsampled to 25 players and 28 sub-attribute columns — creating
a genuinely high dimensional structure where the number of variables exceeds the
number of observations. `set.seed(123)` ensures the same 25 players are selected
on every run, guaranteeing reproducibility.

## 6.3 Run PCA

```{r}
# Center and scale all attributes before PCA
# Centering: subtracts the mean so all attributes have mean zero
# Scaling: divides by SD so all attributes have equal variance
# Essential — without scaling, high-variance attributes would dominate the components
pca_result <- prcomp(pca_data, center = TRUE, scale. = TRUE)

summary(pca_result)
```

## 6.4 Scree Plot — How Many Components Do We Need?

```{r fig.width=8, fig.height=4}
variance_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2) * 100
cumulative_variance <- cumsum(variance_explained)

scree_data <- tibble(
  component = 1:length(variance_explained),
  variance = variance_explained,
  cumulative = cumulative_variance
)

ggplot(scree_data %>% filter(component <= 10), aes(x = component, y = variance)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = cumulative), color = "red", size = 1) +
  geom_point(aes(y = cumulative), color = "red", size = 2) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Scree Plot — Variance Explained by Each Principal Component",
    x = "Principal Component", y = "Variance Explained (%)",
    caption = "Red line shows cumulative variance explained"
  ) +
  theme_minimal()
```

```{r}
cat("Components needed to explain 80% of variance:",
    which(cumulative_variance >= 80)[1], "\n")
cat("Variance explained by first 3 components:",
    round(cumulative_variance[3], 1), "%\n")
```

PC1 alone explains approximately 42% of all variance — the single most important
dimension separating players. The first three components together explain 70.5%
of variance. Six components are needed to explain 80% — meaning the 28 sub-attributes
can be compressed into just 6 independent dimensions without losing much information.

## 6.5 Visualise Players in PCA Space

```{r fig.width=9, fig.height=6}
set.seed(123)
pca_df <- as.data.frame(pca_result$x) %>%
  bind_cols(
    soccer_clean %>%
      drop_na() %>%
      sample_n(25, replace = FALSE) %>%
      mutate(player_type = case_when(
        grepl("CB|LB|RB|LWB|RWB", player_positions) ~ "Defender",
        grepl("CDM|CM|CAM|LM|RM", player_positions)  ~ "Midfielder",
        grepl("ST|CF|LW|RW|LF|RF", player_positions) ~ "Forward",
        TRUE ~ "Other"
      )) %>%
      select(player_type)
  )

ggplot(pca_df %>% filter(player_type != "Other"),
       aes(x = PC1, y = PC2, color = player_type)) +
  geom_point(alpha = 0.7, size = 3) +
  labs(
    title = "Players in PCA Space — PC1 vs PC2",
    x = "PC1 — Technical Attacking Ability",
    y = "PC2 — Defensive & Physical Ability",
    color = "Player Type"
  ) +
  theme_minimal()
```

## 6.6 Interpret the Principal Components

```{r}
loadings <- as.data.frame(pca_result$rotation[, 1:3])
loadings$attribute <- rownames(loadings)

# PC1 — top 10 attributes by absolute loading
loadings %>%
  select(attribute, PC1) %>%
  arrange(desc(abs(PC1))) %>%
  head(10)
```

```{r}
# PC2 — top 10 attributes by absolute loading
loadings %>%
  select(attribute, PC2) %>%
  arrange(desc(abs(PC2))) %>%
  head(10)
```

**PC1 — Technical Attacking Ability**
The attributes with the highest loadings on PC1 are all attacking and technical:
skill_dribbling, skill_ball_control, power_long_shots, mentality_vision,
mentality_positioning, skill_curve, attacking_finishing. PC1 measures how
technically gifted and attack-minded a player is. Players who score high on PC1
are technically elite attackers. Players who score low are more defensive.

**PC2 — Defensive & Physical Ability**
The attributes with the highest loadings on PC2 are all defensive and physical:
mentality_interceptions (-0.370), defending_standing_tackle (-0.361),
defending_sliding_tackle (-0.350), mentality_aggression (-0.341),
power_strength (-0.253). PC2 measures how defensively strong and physically
powerful a player is.

**Football interpretation:**
The two most fundamental dimensions separating professional footballers are:
1. How technically gifted and attack-minded you are — PC1
2. How defensively strong and physically powerful you are — PC2

This mathematically confirms the attacker-defender divide observed throughout
the EDA — it is the most fundamental structural pattern in the data.

---

# 7. Summary

**What this notebook establishes:**

- FIFA 21 scores are not arbitrary — they are position-weighted composites of
  ~30 sub-attributes anchored to real world athletic benchmarks
- Data cleaning followed a comprehensive systematic 8-step framework with clear
  justifications for every decision
- EDA reveals clear position-based structure — distinct attacker, midfielder,
  and defender profiles exist
- Multiple linear regression explains 44.2% of wage variation. Dribbling is the
  strongest skill predictor. Defenders and Forwards earn ~50-52% more than
  Midfielders with identical skills on the original wage scale. The model shows
  assumption violations consistent with the bimodal wage structure.
- PCA on a high dimensional sample (25 players, 28 attributes) reveals two
  fundamental skill dimensions: **technical attacking ability** (PC1, ~42% of
  variance) and **defensive/physical ability** (PC2, ~19% of variance).
  Six components explain 80% of all variance.

**Potential next steps:**
- Model the two wage tiers separately to address the bimodal structure
- Cluster players into archetypes using PCA components
- Explore more complex modelling techniques
